---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
    fig_width: 10
    fig_height: 6
---

# Impact of major storms and weather events in the United States

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This report involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. It tries to address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing
First we load the storm data into `StormData`

```{r}
dataPath <- "~/Downloads/"
StormData <- read.csv(bzfile(paste(dataPath, "StormData.csv.bz2", sep = "")), header = TRUE, stringsAsFactors = FALSE)
```

Then we extract only interested columns: `EVTYPE`, `FATALITIES`, `INJURIES`, `PROPDMG`, `PROPDMGEXP`, `CROPDMG`, `CROPDMGEXP`

```{r}
ExtractedData <- StormData[, c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
```

Let's examine variables `PROPDMGEXP` and `CROPDMGEXP`

```{r}
table(ExtractedData$PROPDMGEXP)
```

```{r}
table(ExtractedData$CROPDMGEXP)
```

Let's convert exponent of `PROPDMGEXP` and `CROPDMGEXP` into numeric values

```{r}
ExtractedData$PROPDMGEXP[ExtractedData$PROPDMGEXP == "" | ExtractedData$PROPDMGEXP == "?" | ExtractedData$PROPDMGEXP == "+" | ExtractedData$PROPDMGEXP == "-"] <- 0
ExtractedData$PROPDMGEXP[ExtractedData$PROPDMGEXP == "H" | ExtractedData$PROPDMGEXP == "h"] <- 2
ExtractedData$PROPDMGEXP[ExtractedData$PROPDMGEXP == "K" | ExtractedData$PROPDMGEXP == "k"] <- 3
ExtractedData$PROPDMGEXP[ExtractedData$PROPDMGEXP == "M" | ExtractedData$PROPDMGEXP == "m"] <- 6
ExtractedData$PROPDMGEXP[ExtractedData$PROPDMGEXP == "B" | ExtractedData$PROPDMGEXP == "b"] <- 9

ExtractedData$PROPDMGEXP <- as.numeric(ExtractedData$PROPDMGEXP)
```

```{r}
ExtractedData$CROPDMGEXP[ExtractedData$CROPDMGEXP == "" | ExtractedData$CROPDMGEXP == "?" | ExtractedData$CROPDMGEXP == "+" | ExtractedData$CROPDMGEXP == "-"] <- 0
ExtractedData$CROPDMGEXP[ExtractedData$CROPDMGEXP == "H" | ExtractedData$CROPDMGEXP == "h"] <- 2
ExtractedData$CROPDMGEXP[ExtractedData$CROPDMGEXP == "K" | ExtractedData$CROPDMGEXP == "k"] <- 3
ExtractedData$CROPDMGEXP[ExtractedData$CROPDMGEXP == "M" | ExtractedData$CROPDMGEXP == "m"] <- 6
ExtractedData$CROPDMGEXP[ExtractedData$CROPDMGEXP == "B" | ExtractedData$CROPDMGEXP == "b"] <- 9

ExtractedData$CROPDMGEXP <- as.numeric(ExtractedData$CROPDMGEXP)
```

We are now ready to compute the "real" values of `PROPDMG` and `CROPDMG`

```{r}
ExtractedData$PROPDMG <- ExtractedData$PROPDMG * (10 ^ ExtractedData$PROPDMGEXP)
ExtractedData$CROPDMG <- ExtractedData$CROPDMG * (10 ^ ExtractedData$CROPDMGEXP)
```

Finally we will aggregate data (`FATALITIES`, `INJURIES`, `PROPDMG`, and `CROPDMG`) by `EVTYPE`

```{r}
FATALITIES <- aggregate(FATALITIES ~ EVTYPE, data = ExtractedData, FUN = sum)
INJURIES <- aggregate(INJURIES ~ EVTYPE, data = ExtractedData, FUN = sum)
PROPDMG <- aggregate(PROPDMG ~ EVTYPE, data = ExtractedData, FUN = sum)
CROPDMG <- aggregate(CROPDMG ~ EVTYPE, data = ExtractedData, FUN = sum)
```

## Results

### Question 1: Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?

Let's take 10 most harmful types of events

```{r}
FATALITIES10 <- FATALITIES[order(FATALITIES$FATALITIES, decreasing = TRUE), ][1:10, ]
INJURIES10 <- INJURIES[order(INJURIES$INJURIES, decreasing = TRUE), ][1:10, ]
```

Top 10 types of events for fatalities

```{r}
FATALITIES10
```

Top 10 types of events for injuries

```{r}
INJURIES10
```

```{r}
par(mfrow = c(1, 2))
barplot(FATALITIES10$FATALITIES, las = 3, names.arg = FATALITIES10$EVTYPE, main = "Top 10 types of weather events for fatalities", ylab = "Number of fatalities", col = "red")
barplot(INJURIES10$INJURIES, las = 3, names.arg = INJURIES10$EVTYPE, main = "Top 10 types of weather events for injuries", ylab = "Number of injuries", col = "red")
```

**"TORNADO" is the most harmful to population health.**

### Question 2: Across the United States, which types of events have the greatest economic consequences?

Let's take 10 types of events with greatest economic consequences

```{r}
PROPDMG10 <- PROPDMG[order(PROPDMG$PROPDMG, decreasing = TRUE), ][1:10, ]
CROPDMG10 <- CROPDMG[order(CROPDMG$CROPDMG, decreasing = TRUE), ][1:10, ]
```

Top 10 types of events for property damages

```{r}
PROPDMG10
```

Top 10 types of events for crop damages

```{r}
CROPDMG10
```

```{r}
par(mfrow = c(1, 2))
barplot(PROPDMG10$PROPDMG / (10 ^ 9), las = 3, names.arg = PROPDMG10$EVTYPE, main = "Top 10 weather events types for property damages", ylab = "Cost of damages ($ billions)", col = "red")
barplot(CROPDMG10$CROPDMG / (10 ^ 9), las = 3, names.arg = CROPDMG10$EVTYPE, main = "Top 10 weather events types for crop damages", ylab = "Cost of damages ($ billions)", col = "red")
```

**"FLOOD" caused the greatest damage to properties. "DROUGHT" caused the greatest damages to crops.**